Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-18 opsem.tex
edits upto Sec. 5

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), and
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
supported nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).

@@

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward. The semantics assumes that  records in
$\stg$ can be uniquely identified via their $\idf$ field, and
enforces this property wherever necessary. Certain hidden fields are
treated specially by the operational semantics, and they are hidden
from the surface language. These include a $\txnf$ field that tracks
the identifier of the transaction that last updated the record, and a
$\delf$ field that flags deleted records in $\stl$.  For a set $S$ of
records, we define $\dom(S)$ as the set of unique ids of all records
in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, most popular databases enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, whose $\delf$ field is set to \C{true}. When the transaction
commits, its local database is atomically \emph{flushed} to the global
database, committing these heretofore uncommitted records. The flush
operation ($\gg$) is defined as follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, provided the update is
not a deletion ($\neg r.\delf$).  Besides the commit, flush also helps
a transaction read its own writes. Intuitively, the result of a read
operation inside a transaction must be computed on the database
resulting from flushing the current local state ($\stl$) to the global
state ($\stg$). The abstract machine of Fig.~\ref{fig:txnimp},
however, does not let a transaction read its own writes. This
simplifies the semantics, without losing any generality, since
substituting $\stl\gg\stg$ for $\stg$ at select places in the
reduction rules effectively allows reads of uncommitted transaction
writes to be realized.

@@

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction comes
tagged with an isolation specification $\I$, which has two components
$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
that the transaction can witness, during its execution ($\I_e$), and
when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
predicates over the (current) transaction-local database state
($\stl$), the state ($\stg$) of the global database when the
transaction last took a step, and the current state ($\stg'$) of the
global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in \S\ref{sec:motivation}, an SI transaction on
PostgreSQL defines $\I$ as follows:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($[\bar{f}=\bar{e}]$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), 
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).

@@

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward, e.g., by having the machine manipulate a
set of sets, one for each table.  The semantics assumes that records
in $\stg$ can be uniquely identified via their $\idf$ field, and
enforces this property wherever necessary. Certain hidden fields are
treated specially by the operational semantics, and are hidden
from the surface language. These include a $\txnf$ field that tracks
the identifier of the transaction that last updated the record, and a
$\delf$ field that flags deleted records in $\stl$.  For a set $S$ of
records, we define $\dom(S)$ as the set of unique ids of all records
in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, most popular databases enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, whose $\delf$ field is set to \C{true}. When the transaction
commits, its local database is atomically \emph{flushed} to the global
database, committing these heretofore uncommitted records. The flush
operation ($\rhd$) is defined as follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\rhd\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\rhd\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, provided the update is
not a deletion ($\neg r.\delf$).  Besides the commit, flush also helps
a transaction read its own writes. Intuitively, the result of a read
operation inside a transaction must be computed on the database
resulting from flushing the current local state ($\stl$) to the global
state ($\stg$). The abstract machine of Fig.~\ref{fig:txnimp},
however, does not let a transaction read its own writes. This
simplifies the semantics, without losing any generality, since
substituting $\stl\rhd\stg$ for $\stg$ at select places in the
reduction rules effectively allows reads of uncommitted transaction
writes to be realized.

@@

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction comes
tagged with an \emph{isolation specification} $\I$, which has two components
$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
that the transaction can witness, during its execution ($\I_e$), and
when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
predicates over the (current) transaction-local database state
($\stl$), the state ($\stg$) of the global database when the
transaction last took a step, and the current state ($\stg'$) of the
global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in \S\ref{sec:motivation}, an SI transaction on
PostgreSQL defines $\I$ as follows:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
This definition dictates that no change to the global database state
can be visible to an SI transaction while it executes ($\I_e$), and
there should be no concurrent updates to records written by the
transaction by other concurrently executing ones ($\I_c$).
To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

### 2017-10-18 reasoning.tex
edits upto Sec. 5

 \begin{smathpar}
\forall \stl, \stg, \stg', \stg_1. Q(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg') \Rightarrow Q(\stl, \stg')
\end{smathpar}

In the above condition, we use a separate notation ($\stg_1$) for the
global database state that the transaction witnessed during its last
step. $\phi$ does not constrain this global database state, but the
stability of $\I$ ensures that any database state constrained by
$\phi$ must also obey $\I$. We can now replace the transitive closure
in the above condition with the single step relation $R$, backed by
the inductive argument for stability. Pushing the universal quantifier
for $\stg_1$ inside the antecedent and separating the predicates
involving $\stg_1$, we can rewrite the above condition into the form
shown below, by defining a new ternary relation $\R$, which can be
thought of as $R$ modulo $\I$:

\begin{smathpar}
\begin{array}{lcl}
\stable(\R, Q) & \Leftrightarrow & \forall \stl,\stg,\stg'. Q(\stl,\stg) \wedge \R(\stl,\stg,\stg') \Rightarrow Q(\stl,\stg')\\
\R(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}

The above condition often significantly simplifies the form of $\R$
irrespective of $R$. For example, when a transaction is executed
against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R$ will
be the identity function, since if $\stg \neq \stg'$, there does not
exist any $\stg_1$ which satisfies the constraint for
$\R_e(\stl,\stg,\stg')$.

>>>>

 \begin{smathpar}
   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
   \Rightarrow \phi(\stl, \stg')
\end{smathpar}

In the above condition, we use a separate notation ($\stg_1$) for the
global database state that the transaction witnessed during its last
step. $\phi$ does not constrain this global database state, but the
stability of $\I$ ensures that any database state constrained by $\phi$
must also obey $\I$. We can now replace the transitive closure in the
above condition with the single step relation $R$, backed by the
inductive argument for stability. Pushing the universal quantifier for
$\stg_1$ inside the antecedent and separating the predicates involving
$\stg_1$, we can rewrite the condition in the form below, by defining
a new ternary relation $R \backslash \I$, which can be thought of as
$R$ modulo $\I$:

\begin{smathpar}
\begin{array}{lcl}
\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}

The above condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since if $\stg
\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.
